---
title: "SensorOverlord Walkthrough"
output: 
  rmarkdown::html_vignette: default
  word_document: default
vignette: >
  %\VignetteIndexEntry{SensorOverlord Walkthrough}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(sensorOverlord)
```


## Purpose and Audience

This short tutorial is intended for microscopists who have:

* A basic knowledge of R,

* access to ratiometric images in their experimental system of interest, and

* an interest in making accurate, quantitative measurements of $E_{GSH}$, pH, or 
the concentration of a ligand.

## Obtaining Pre-requisities

### Installing the sensorOverlord Package

The sensorOverlord package is available on GitHub. You can install it in R via the 
`devtools` package. To install the `devtools` package, run `install.packages("devtools")`.

To install the most recent version of the sensorOverlord package from GitHub:

```R
devtools::install_github("apfeldlab/sensoroverlord")
```

You can confirm your installation of `sensorOverlord` with `packageVersion`:

```{R}
packageVersion("sensorOverlord")
```

### Obtaining Sensor Biochemical and Biophysical Parameters

#### 1. Biochemical parameters

To use the SensorOverlord framework, you need to have access to a midpoint potential
for your sensor. 

For redox sensors, the midpoint potential is $E_0$. For pH sensors, the midpoint
potential is $pKa$. For ligand-binding sensors, the midpoint is $pKd$. 

The value of these midpoint potentials are almost always reported in the paper 
in which the sensor of interest was first published. 

Note that, for ligand-binding sensors, papers often report $Kd$. $pKd = -log_{10}(Kd)$ so,
to convert $Kd$ to $pKd$, use the `log10` function:

```R
kD <- 5*10e-6
pKd <- -log(kD)
```

#### 2. Biophysical parameters

sensorOverlord requires as input:

1. $R_A$: The ratio measurement of the sensor in State A. We also assume that State A
is the state in which you have the _lowest_ ratio measurement, so this is sometimes also called
$R_{min}$.
2. $R_B$: The ratio measurement of the sensor in State B. Sometimes also called $R_{max}$.
3. $\delta_{\lambda2}$: The distribution of the dynamic range in the second wavelength. This can be derived from the dynamic range of the sensor in both states.

You can obtain these parameters in one of three ways:

1. Upload a digitized spectra. `sensorOverlord` can programatically obtain these
input values from a digitized spectra.

2. Use a provided spectra. We have created a MongoDB database containing many the spectra of many
published redox, pH, and ligand-binding sensors.

3. Provide the parameters directly. These parameters can be measured empirically, 
as explained in [Romero 2014](https://www.nature.com/articles/ncomms6020).


**Option 1: Upload a digitized spectra**

If you have excitation-emission spectra data for your each of your sensor's two
states, you can upload those into R. 

For example, you could have a .csv file with three columns:

* lambda: The wavelength of the excitation or emission.
* values_A: The intensity value of the sensor's output in state A.
* values_B: The intensity value of the sensor's output in state B.

Oftentimes with digitized spectra, the $\lambda$ values recorded in states A
and B are not identical. In that case, you can use a .csv file with four columns:

* lambda_A: The wavelengths recorded in state A.
* values_A: The intensity value of the sensor's output in state A.
* lambda_B: The wavelengths recorded in state B.
* values_B: The intensity value of the sensor's output in state B.

```R
spectra <- read.csv("my_spectra.csv")
```

Once you have uploaded your spectra, you can use the `sensorOverlord::spectraMatrixFromValues` function
to create a `Spectra` object:

```R
spectra_object <- spectraMatrixFromValues(
  lambdas_minimum = spectra$lambda,
  values_minimum = spectra$values_A,
  lambdas_maximum = spectra$lambda,
  values_maximum = spectra$values_B
)
```

or

```R
spectra_object <- spectraMatrixFromValues(
  lambdas_minimum = spectra$lambda_A,
  values_minimum = spectra$values_A,
  lambdas_maximum = spectra$lambda_B,
  values_maximum = spectra$values_B
)
```

You can use the `newSensorFromSpectra()` function to turn the 
spectra object into a sensor object at a certain ratiometric wavelength pair:

```R
sensor_object <- newSensorFromSpectra(spectra_object, 
  c(510, 520), c(455, 465))
```

That sensor object will now contain estimates for the three required parameters:

```R
Ra <- sensor_object@Rmin
Rb <- sensor_object@Rmax
delta2 <- sensor_object@delta
```

**Option 2: Use a Provided Spectra**

You can access all sensor information from the database with `sensorOverlord::getDb()$find()`,
which accepts JSON search queries. For more information, see the "Accessing and Manipulating the Database"
vignette. For example:

```{R}
deGFP1_information <- getDb()$find('{"sensor_name" : "deGFP1"}')
print(deGFP1_information$sensor_readout)
print(deGFP1_information$sensor_midpoint)
print(deGFP1_information$lambda1_recommended)
print(deGFP1_information$lambda2_recommended)
```

The actual spectra information is in a list form, so you can use the `apply` 
and `unlist` functions to obtain the spectra:

```{R}
library(data.table)
deGFP1_spectra_information <- data.table(
  apply(
    deGFP1_information[c("lambda_min", "values_min", "lambda_max", "values_max")], 
    2, function(x) unlist(x))
  )
```

Then, you can follow the same instructions as if you had provided the spectra yourself:
create a spectra object, then a sensor object at a specific wavelength pair, and then
extract the relevant parameters:

```{R}
deGFP_spectra <- spectraMatrixFromValues(deGFP1_spectra_information$lambda_min,
                                                         deGFP1_spectra_information$values_min,
                                                         deGFP1_spectra_information$lambda_max,
                                                         deGFP1_spectra_information$values_max)

sensor_object <- newSensorFromSpectra(deGFP_spectra,
                                     c(510, 520), c(455, 465))

Ra <- sensor_object@Rmin
Rb <- sensor_object@Rmax
delta2 <- sensor_object@delta

print(Ra)
print(Rb)
print(delta2)
```

**Option 3: Provide Parameters Directly**

If you conducted the appropriate microscopy experiments, you may have also 
obtained exact values of $R_a$, $R_b$, and $\delta_{\lambda2}$.

If that's the case, you can use them to directly make a sensor object:

```{R}
R_reduced <- 0.6
R_oxidized <- 5.2
delta2 <- 0.171

sensor_object <- new(
            "Sensor",
            Rmax = R_reduced,
            Rmin = R_oxidized,
            delta = delta2
        )
```

### Estimating Microscopy Error

In order to use the SensorOverlord framework, you need to have an estimate of
the variability of your measurement of $R$ in your system of interest.

There are various ways to accomplish this, as we detail in the Supplementary Information of the associated manuscript.

For example, you can simply plot the $R$ values that you observed in a previous
experiment in which you expected the $R$ value to stay constant over some period
of time. Then, you can estimate your error in $R$ by looking at the standard error
over that period.

In C. elegans, we find that the error is constant, but relative to the $R$ value.

SensorOverlord can accommodate any error model that you find in your system. 
For example:

```{R}
# Relative error of 2.8%, or 0.028*R
error_relative <- function(x) x*0.028

# Absolute error of +/- 0.028 
error_absolute <- function(x) 0.028

# Quadratic error
error_other <- function(x) 0.0001 * x^2 + 0.028*x + 0.5
```

For more information about incorporating different error models, see the "Customizing Error Model"
vignette. For now, this tutorial will assume that you are using a _relative_ error.

## Running the SensorOverlord model

After completing the pre-requisites, you should have access to a sensor object
and an error model. E.g.

```{R}
sensor_object <- new(
            "Sensor",
            Rmax = R_reduced,
            Rmin = R_oxidized,
            delta = delta2
        )

my_error <- function(x) x*0.028
```
`

First, you'll need to turn your generic sensor object into a specific redox, pH, or ligand-binding sensor. For example:

```{R}
redox_sensor_object <-
  new(
    "redoxSensor",
    sensor_object,
    e0 = -265
  )
```

Then, you can use the `error_df()` method, defined on redoxSensor, pHSensor, and ligandSensor objects, to get a table of the inaccuracy of the sensor at different values of redox potential, pH, or p[Ligand].

```{R}
redox_sensor_error <- error_df(
  redox_sensor_object,
  c(0.028),
  Emin = -300,
  Emax = -150
)

summary(redox_sensor_error)
```

```{R}
library(ggplot2)
ggplot(redox_sensor_error, aes(x = E, y = Error)) +
  geom_line() +
  ylim(c(0, 10))
```

For more information, see `?error_df,redoxSensor-method`, or the method for your type of sensor.

You can export that table directly, or use it for further analysis. 

`sensorOverlord` can also compute the ranges of redox potential, pH, or ligand concentration that a sensor can measure to different accuracy thresholds:


```{R}
# The redox potentials this sensor can measure, with an error in R of 
# 2% or 3%, to an accuracy of 1mV or 2mV
accurate_ranges <- ranges_df(redox_sensor_object,
                                  inaccuracies = c(0.02, 0.03),
                                  thresholds = c(1, 2))

accurate_ranges
```

`sensorOverlord` also provides some default plotting methods that work for all sensor objects.

`plotProperty` will plot $R$ vs the property of interest For example, for redox
sensors, this will be $R$ vs $E_{GSH}$, the fraction of sensors in the oxidized state.

```{R}
plotProperty(redox_sensor_object)
```

`rangePlot` will plot the range of values a sensor can measure to a certain accuracy.

```{R}
rangePlot(redox_sensor_object, ranges = accurate_ranges)
```

In the plot above, notice how the range of values the sensor can measure accurately
expands when the error in R is 2% ("Sensor_0.02") versus when the error in 3% ("Sensor_0.03").

For more information on plotting methods, see the "Sensor Methods" section in [the documentation reference](https://apfeldlab.github.io/SensorOverlord/reference/index.html).
